### 2021-12-13 documents/floats/pwaireg.tex
changed reconstruction of 'to come'

\begin{table}[h]
\centering
\caption[Regular \PWai verbs]{Regular \qu{to fall} (\gl{s_a_}) and \qu{to sleep} (\gl{s_p_}) in \PWai \parencites[150]{howard2001wrought}[189, 190, 196]{hixkaryanaderby1985}[30]{waiwaihawkins1998}[209, 211]{hawkins1953waiwai}}
\label{tab:pwaireg}
\begin{tabular}[t]{@{}lllllll@{}}
\mytoprule
{} &           \PWai &                   &           \hixka &                 &            \waiwai &                    \\
{} &    \qu{to fall} &     \qu{to sleep} &     \qu{to fall} &   \qu{to sleep} &       \qu{to fall} &      \qu{to sleep} \\
\mymidrule
\gl{1}   &  \rc{k-eɸurka-} &   \rc{kɨ-wɨnɨkɨ-} &  \obj{k-ehurka-} &  \obj{kɨ-nɨkɨ-} &    \obj{k-eɸɨrka-} &   \obj{kɨ-wɨnɨkɨ-} \\
\gl{2}   &  \rc{m-eɸurka-} &    \rc{o-wɨnɨkɨ-} &  \obj{m-ehurka-} &  \obj{o-wnɨkɨ-} &    \obj{m-eɸɨrka-} &   \obj{mɨ-wɨnɨkɨ-} \\
\gl{1+2} &  \rc{t-eɸurka-} &  \rc{tɨt-wɨnɨkɨ-} &  \obj{t-ehurka-} &  \obj{tɨ-nɨkɨ-} &  \obj{t͡ʃ-eɸɨrka-} &  \obj{tɨt-wɨnɨkɨ-} \\
\gl{3}   &  \rc{ɲ-eɸurka-} &   \rc{nɨ-wɨnɨkɨ-} &  \obj{ɲ-ehurka-} &  \obj{nɨ-nɨkɨ-} &    \obj{ɲ-eɸɨrka-} &   \obj{nɨ-wɨnɨkɨ-} \\
\mybottomrule
\end{tabular}
\end{table}

>>>>

\begin{table}[h]
\centering
\caption[Regular \PWai verbs]{Regular \qu{to fall} (\gl{s_a_}) and \qu{to sleep} (\gl{s_p_}) in \PWai \parencites[30]{waiwaihawkins1998}[150]{howard2001wrought}[189, 190, 196]{hixkaryanaderby1985}[209, 211]{hawkins1953waiwai}}
\label{tab:pwaireg}
\begin{tabular}[t]{@{}lllllll@{}}
\mytoprule
{} &           \PWai &                   &           \hixka &                 &            \waiwai &                    \\
{} &    \qu{to fall} &     \qu{to sleep} &     \qu{to fall} &   \qu{to sleep} &       \qu{to fall} &      \qu{to sleep} \\
\mymidrule
\gl{1}   &  \rc{k-eɸurka-} &   \rc{kɨ-wɨnɨkɨ-} &  \obj{k-ehurka-} &  \obj{kɨ-nɨkɨ-} &    \obj{k-eɸɨrka-} &   \obj{kɨ-wɨnɨkɨ-} \\
\gl{2}   &  \rc{m-eɸurka-} &    \rc{o-wɨnɨkɨ-} &  \obj{m-ehurka-} &  \obj{o-wnɨkɨ-} &    \obj{m-eɸɨrka-} &   \obj{mɨ-wɨnɨkɨ-} \\
\gl{1+2} &  \rc{t-eɸurka-} &  \rc{tɨt-wɨnɨkɨ-} &  \obj{t-ehurka-} &  \obj{tɨ-nɨkɨ-} &  \obj{t͡ʃ-eɸɨrka-} &  \obj{tɨt-wɨnɨkɨ-} \\
\gl{3}   &  \rc{ɲ-eɸurka-} &   \rc{nɨ-wɨnɨkɨ-} &  \obj{ɲ-ehurka-} &  \obj{nɨ-nɨkɨ-} &    \obj{ɲ-eɸɨrka-} &   \obj{nɨ-wɨnɨkɨ-} \\
\mybottomrule
\end{tabular}
\end{table}

### 2021-12-13 documents/sections/verbs.tex
changed reconstruction of 'to come'

\subsubsection{\rc{(ət-)jəpɨ} \qu{to come}}
\label{sec:come}
This verb can be reconstructed as \PC \rc{(ət-)jəpɨ}, morphologically segmentable into an optional detransitivizing prefix and a root \rc{jəpɨ}.
As discussed in \cref{sec:pekodian}, innovative \rc{k-} was introduced on the \ikpeng and \bakairi reflexes of \rc{ətjəpɨ}, but not on the \arara reflex of \rc{jəpɨ}.
Reflexes of \rc{ətjəpɨ} (\trio) and of \rc{ətjəpɨ}/\rc{jəpɨ} (\akuriyo) resisted the introduction of \PTir \rc{t-}.
\carijo \obj{ehɨ} shows innovative \obj{j-}, rather than conservative \obj{w-} \exref{car-18}.
No \yukpa reflex of this verb is attested, and it was fully replaced in \PWai by \rc{omokɨ} \qu{to come}.

\ex<car-18>\carijo \parencite[][102]{guerrero2019carijo}\\
\begingl
\gla əji-wa-e j-eh-ɨ//
\glb \gl{2}-search-\gl{sup} \gl{1}-come-\gl{pfv}//
\glft \qu{I came looking for you.}//
\endgl
\xe

@@

This reconstruction disagrees with \posscite[30]{gildea2007greenberg} \rc{ətepɨ}, due to the comparative evidence in \cref{tab:come}.
Crucially, most reflexes do not reflect the \rc{ət} part of \posscite{gildea2007greenberg} reconstruction, and many ostensibly go back to \rc{əpɨ}, \rc{jepɨ}, or \rc{jəpɨ}, rather than \rc{epɨ}.

Only the Pemongan languages and \kaxui point to an \rc{j}-initial root, and \kaxui \obj{johɨ} is very rare in contrast to more frequent \obj{o(o)hɨ}.
It only occurs in the third person form of the Progressive, meaning that the \obj{j} may be a reflex of third person \rc{i-}.
However, regular \obj{o}-initial \kaxui verbs have no third person prefix \pcref{tab:kaxprog}, while \obj{i-} occurs with C-initial \obj{to[mo]} \qu{to go}.
Even if \qu{to come} once shared this \obj{i-}, the \obj{j} is best analyzed as part of the root in \kaxui, allowing the reconstruction of \PC \rc{jəpɨ}.
%
\input{floats/kaxprog}%
%\pex<kaxprog>\kaxui \perscomm{Spike Gildea}
%\a<kax-82>
%\begingl
%\gla johɨ-rɨ//
%\glb \gl{3}.come-\gl{prog}//
%\glft \qu{S/he is coming.}//
%\endgl
%\a<kax-54>
%\begingl
%\gla o-w-ohɨ-rɨ//
%\glb \gl{2}-\gl{s_a_}-come-\gl{prog}//
%\glft \qu{You are coming.}//
%\endgl
%\a<kax-77>
%\begingl
%\gla i-nkɨ-rɨ//
%\glb \gl{3}-sleep-\gl{prog}//
%\glft \qu{S/he is sleeping.}//
%\endgl
%\a<kax-83>
%\begingl
%\gla Ø-osone-rɨ//
%\glb \gl{3}-dream-\gl{prog}//
%\glft \qu{S/he is dreaming.}//
%\endgl
%\xe
%
The morphologically complex (non-Pemongan) forms corresponding to \posscite{gildea2007greenberg} \rc{ətepɨ} show no segmental trace of \rc{j}, but the \obj{i} in \akawaio \obj{əsipɨ} likely reflects the sequence \rc{jə}.
This analysis is supported by bare \rc{jəpɨ} becoming \obj{ipɨ} in closely related \macushi.

The many forms seemingly reflecting \rc{əpɨ} and \rc{epɨ} rather than \obj{jəpɨ} are distributed unsystematically in the family, sometimes even co-occurring in the same language.
A unifying account requires the root \rc{jəpɨ} to undergo two main sound changes: \begin{inlinelist}
 \item \rc{j}-loss
 \item \rc{ə}-umlaut after \rc{j}
 \end{inlinelist}.
Both sound changes are found in other contexts throughout the family \parencite{meira2010origin}, but they appear to have applied inconsistently to this verb, and not always in the same order.
Forms like \kalina \obj{opɨ} can only be explained if \rc{j} was lost before the umlauting of \rc{ə} to \rc{e}, which would have been triggered by \rc{j}.
On the other hand, forms like \maqui \obj{ehə} must be the result of \phonl{\rc{ə}}{\rc{e}}{\rc{j}}, with subsequent loss of \rc{j}.
The \akuriyo form \obj{eepɨ} looks like a reflex thereof as well, but the length is unexpected, and is analyzed by \textcite[]{meira1998proto} as a coalescence of \rc{e} and \rc{əe}.

%Other languages -- across the family -- have forms reflecting the same root without the initial consonant, i.e.\ \rc{əpɨ}: \kalina, \panare, \yawarana, and \uxc.\footnote{\obj{e} is the regular outcome of \rc{ə} in \uxc; \rc{e} became \obj{i} \parencite[176]{meira2005southern}.}
%Also found across the family are forms reflecting \rc{epɨ}, in \kaxui, \arara, \trio, \carijo, \maqui, and \mapoyo.

While a root \rc{jəpɨ}, the two sound changes, and the optional addition of \rc{ət-} do account for the majority of the forms in \cref{tab:come},\footnote{
	Apart from aforementioned \akuriyo \obj{eepɨ}, another diachronically irregular form is \apalai \obj{oepɨ}, where one would expect \rc{ət-epɨ} to yield \obj{os-epɨ} \parencite[506]{meira2010origin}.
	Similarly, while \obj{oepɨ} would be a regular outcome of a hypothetical \rc{ə-jəpɨ}, the \envr{}{C} allomorph of the detransitivizer is \obj{e-} in \apalai.
	One possibility is that the form is due to borrowing from \trio, which has lost intervocalic \rc{t} to create \obj{əepɨ}.
	Alternatively, \apalai \obj{oepɨ} could be a fossilized loan from \wayana, which has replaced its reflex of \rc{ətjəpɨ}, but where intervocalic \rc{t} was also regularly lost \parencite[63]{wayanatavares2005}.}
the distribution within the family is rather chaotic.
In addition to the seemingly unordered distribution of \rc{əpɨ} and \rc{epɨ}, forms with and without \rc{ət-} can be found within the same language, usually conditioned by different prefixes.
This was briefly discussed in \cref{sec:pekodian} for \arara (and \PPek) and in \cref{sec:taranoan} for \trio (and \PTar).
To illustrate, the \trio \setone paradigm shows a reflex of \rc{ətepɨ} (< \rc{ətjəpɨ}) for first, but of \rc{epɨ} (< \rc{jəpɨ}) for the other persons \exref{tricome}.%
\footnote{While the \gl{1+2} form is a regular outcome of \rc{kɨt-epɨ}, the second person form is mysterious \parencite[115]{meira1998proto}.}
It should be noted that forms with and without \rc{ət-} in different languages are not triggered by the same person values.

\ex<tricome> \trio \parencite[294]{triomeira1999}\\
\begin{tabular}[t]{@{}ll@{}}
	\gl{1} & \obj{w-əepɨ} \\
	\gl{2} &  \obj{mən-epɨ} \\ 
	\gl{1+2} &  \obj{ke-epɨ} \\
	\gl{3} &  \obj{n-epɨ} \\
\end{tabular}
\xe

The interpretation of the \rc{ət} part as a detransitivizer is based on its form and its paradigmatically conditioned occurrence in some languages.
Although the combination of a detransitivizer and an intransitive verb seems semantically illogical, some historical \gl{s_p_} verbs are attested as adding the detransitivizer to become \gl{s_a_} verbs.
For example, the \PC \gl{s_p_} verb \rc{wɨnɨkɨ} \qu{to sleep} becomes \trio \obj{əənɨkɨ} \parencite[252]{triomeira1999} and \kalina \obj{əʔnɨkɨ} \parencite[429]{courtz2008carib}, both \gl{s_a_}.
Also, \waiwai \qu{go to sleep} can be \obj{wɨnɨk} \parencite[30]{waiwaihawkins1998} or \obj{et-wɨnɨk} \parencite[204]{hawkins1953waiwai}.
The parallels to \qu{to sleep} end here, since bare \rc{jəpɨ} \qu{to come} apparently already was an \gl{s_a_} verb, as evidenced the class membership of its reflexes in \kaxui, \kalina, \arara, \trio, and \panare \exref{pan-128}.

\ex<pan-128>\panare \parencite[][65]{panarepayne2013}\\
\begingl
\gla ju-w-əəpɨ-n ka=m kanoʔ//
\glb \gl{3}-\gl{s_a_}-come-\gl{nspec} \gl{q}=\gl{2}.\gl{aux} rain//
\glft \qu{Do you think it is gonna rain?}//
\endgl
\xe

Summing up, this verb is highly irregular, both from a synchronic and diachronic perspective.
The scenario suggested here involves reflexes of the detransitivizer \rc{ət(e)-} being optionally added to an \gl{s_a_} verb root \rc{jəpɨ}, which further underwent umlaut and loss of \rc{j}, but in no systematic manner, resulting in the chaotic picture in \cref{tab:come}.

>>>>

\subsubsection{\rc{(ət-)epɨ} \qu{to come}}
\label{sec:come}
Innovative \rc{k-} was introduced on the \ikpeng and \bakairi reflexes of \rc{ət-epɨ}, but not on the \arara reflex of \rc{epɨ} \pcref{sec:pekodian}.
The reflex of \rc{ət-epɨ} resisted the introduction of \PTir \rc{t-} \pcref{sec:taranoan}.
\carijo \obj{ehɨ} shows innovative \obj{j-}, rather than conservative \obj{w-} \exref{car-18}.
No \yukpa reflex of this verb is attested, and it was fully replaced in \PWai by \rc{omokɨ} \qu{to come}.

\ex<car-18>\carijo \parencite[][102]{guerrero2019carijo}\\
\begingl
\gla əji-wa-e j-eh-ɨ//
\glb \gl{2}-search-\gl{sup} \gl{1}-come-\gl{pfv}//
\glft \qu{I came looking for you.}//
\endgl
\xe
%
\posscite{gildea2007greenberg} reconstruction \rc{ətepɨ} can be segmented into an optional prefix \rc{ət-} and a root \rc{epɨ}, since reflexes can be grouped into those with a reflex of \rc{ət-} and those without \pcref{tab:come}.
As seen in \cref{sec:pekodian,sec:taranoan,tab:comepara}, this division can exist within a single paradigm.

@@

\input{floats/comepara}

Long \rc{ət-epɨ} lost \rc{t} in \trio, \apalai, and \bakairi, but only the first development is due to regular sound changes \parencite[31--32]{meira1998proto}.
\akuriyo reflects \rc{əe} as \obj{ee}, with \obj{eepɨ} being betrayed as not a reflex of \rc{epɨ} by the otherwise unexpected vowel length \parencite[114--115]{meira1998proto}.
Other languages coalesced \rc{əe} to \rc{əə}, with \rc{əəpɨ} being reflected in \uxc \parencite[452]{gildea2012classification}, \kalina, \kaxui, \panare, and \yawarana.
In \kalina \parencite[509--510]{meira2010origin}, \rc{e}-initial transitive verbs detransitivized with \rc{ət-} underwent the same development (also yielding vowel length), while \rc{əte} > \rc{əə} is irregular at least in \kaxui and \panare.

The Pemongan languages and \kaxui point to \rc{jəpɨ} rather than \rc{epɨ}, although \kaxui \obj{johɨ} is very rare in contrast to more frequent \obj{o(o)hɨ}.
It only occurs in the third person form of the Progressive, meaning that the \obj{j} may be a reflex of third person \rc{i-}.
However, regular \obj{o}-initial \kaxui verbs have no third person prefix \pcref{tab:kaxprog}, while \obj{i-} occurs with C-initial \obj{to[mo]} \qu{to go}, suggesting that \obj{j} is indeed part of the root.
Putative \PC \rc{jəpɨ} would yield \rc{epɨ} by the two widespread sound changes of \rc{ə}-umlaut after \rc{j} and \rc{j}-loss \parencite{meira2010origin}, but one would expect evidence of a sequence \rc{jə} in \yawarana, \mapoyo, or \panare \parencite{gildea2010story}.
The \kaxui form \obj{ehɨ} would also not be accounted for, as \rc{ə} was not umlauted after \rc{j} in Parukotoan \parencite[501--502]{meira2010origin}.
%
\input{floats/kaxprog}%

The clear segmentability of \rc{ət-} in combination with its form suggest that it is a detransitivizing prefix.
Although the combination of a detransitivizer and an intransitive verb makes no sense semantically, some historical \gl{s_p_} verbs are attested as adding the detransitivizer to become \gl{s_a_} verbs.
For example, the \PC \gl{s_p_} verb \rc{wɨnɨkɨ} \qu{to sleep} becomes \trio \obj{əənɨkɨ} \parencite[252]{triomeira1999} and \kalina \obj{əʔnɨkɨ} \parencite[429]{courtz2008carib}, both \gl{s_a_}.
\waiwai \qu{to sleep} can be \obj{wɨnɨk} \parencite[30]{waiwaihawkins1998} or \obj{et-wɨnɨk} \parencite[204]{hawkins1953waiwai}.
However, unlike \qu{to sleep}, bare \rc{epɨ} \qu{to come} apparently already was an \gl{s_a_} verb (although its reflexes in languages with split-\gl{s} always co-occur with reflexes of \rc{ət-epɨ}).
